Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-12-09 manuscript.qmd
work through most of Jan's comments

In treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted. Current guidelines also include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.

These CTV-N guidelines rely on anatomically defined lymph node levels (LNLs) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases within these levels. They often recommend extensive irradiation of both sides of the neck. However, the general prevalence of metastasis in a given LNL does not correspond to an individual patient's risk of occult disease in that region, which depends on their specific state of tumor progression. For example, a patient with no clinically detectable nodal disease (cN0) who has a small, clearly lateralized T1 tumor would receive the same contralateral CTV-N as a patient with significant ipsilateral nodal involvement and an advanced tumor crossing the mid-sagittal plane: Both patients receive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].

To better quantify individualized risk of occult disease, we previously developed an intuitive probabilistic hidden Markov model (HMM) [@ludwig_hidden_2021;@ludwig_modelling_2023], originally based on a conceptually similar a Bayesian network model [@pouymayou_bayesian_2019]. However, these models have been limited to predicting ipsilateral nodal involvement. This work extends the model to include contralateral risk predictions, enabling more personalized radiation volume recommendations. By identifying patients with low contralateral risk, the model could guide reductions in contralateral CTV-N, thereby decreasing radiation-induced toxicity and improving quality of life.

@@

```{python}
#| echo: false
#| label: tbl-data-overview
#| tbl-cap: Overview over the five datasets from four different institutions used to train and evaluate our model. Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received neck dissection, the N0 portion of patients, what percentage presented with early T-category, and the prevalence of primary tumor midline extension. For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
from typing import Literal

@@

The datasets enable analysis of correlations between LNL involvement and key risk factors. In @fig-data-strat, we illustrate the prevalence of contralateral LNL involvement, stratified by T-category, the number of ipsilaterally involved LNLs, and whether the tumor extends across the mid-sagittal plane.

@@

1. **Midline Extension:** Tumors extending across the mid-sagittal plane should result in a significantly higher probability of contralateral metastases.
2. **T-Category:** Advanced T-category should correspond to an increased risk of nodal disease. In the hidden Markov model this can be modeled using the expected time of diagnosis, as demonstrated previously for the ipsilateral model [@ludwig_hidden_2021].
3. **Ipsilateral Involvement:** The model should be able to capture the correlation between the extent of ipsi- and contralateral involvement. <!-- TODO: finish -->

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL's may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is healthy or free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to the presence of clinically detected metastases (i.e., occult or macroscopic disease). In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. Each LNL's state is observed via another binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that LNL $v$ was classified as involved. $X_v$ and $Z_v$ are connected through the sensitivity and specificity of the diagnositc modality.

@@

We can evolve the distribution of all possible hidden states from $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$: $\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}$. For later use, we define at this point a matrix $\boldsymbol{\Lambda}$ that collects these distributions for all considered time-steps:

@@

Where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.

@@

@@

The probabiltiy for a patient to present with a diagnosis $\boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:

@@

which represents a single patient's contribution to the overall likelihood function.

Above, $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets. Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.

The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, the parameter space of the unilateral model is:

@@

A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in @sec-unilateral, with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$). However, this method would fail to capture the correlation between ipsilateral and contralateral involvement discussed in @sec-data-strat, particularly the observed increase in contralateral involvement with greater severity of ipsilateral spread.

<!-- This is displayed in @tbl-data-strat and shows how often the contralateral LNLs I, II, III, and IV were involved, given all possible combinations of midline extension, T-category, and ipsilateral LNL III involvement. Unsurprisingly, the prevalence for contralateral involvement is consistently higher when the tumor extends over the mid-sagittal plane or is of later T-category. But it is also more frequent when the ipsilateral side shows more severe involvement, which is here shown via the surrogate LNL III. -->

<!-- Additionally, we could think of an approach to incorporate the primary tumor's mid-sagittal extension as a risk factor. -->

@@

<!-- There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.-->

@@

The matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ completely. However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:

1. **Shared Graph Structure**: Both ipsilateral and contralateral spread are described by the same graph shown in @fig-full-graph.
2. **Symmetric Spread Among LNLs**: The spread among LNLs is assumed to be the same on both sides, reflecting the symmetric structure of the lymphatic system. Consequently, the spread rates between nodes should also be symmetric. This is formalized as:\
  $$
  t_{rv}^\text{c} = t_{rv}^\text{i}
  $$ {#eq-symmetries}
  for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. **Asymmetric Spread from Tumor**: Tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane. When the tumor extends over the midline, its spread rate to the contralateral LNLs should be larger than if the tumor were lateralized, but still smaller than the tumor's spread to the ipsilateral side. To model this, we introduce a mixing parameter $\alpha \in [0,1]$, representing the "degree of asymmetry" effectively mixing these two extremes:\
  $$
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  $$ {#eq-mixing}
  This parameterization results in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$. Despite this additional complexity, these rates collectively account for only $2 \cdot V + 1$ parameters due to their coupling via $\alpha$.

The parameter space now becomes:

@@

Most tumors crossing the midline at the time of diagnosis likely began as lateralized tumors that grew over the midline at a later stage. As a result, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ applies only to a subset of time-steps.

To account for this, we model the tumor's extension over the mid-sagittal plane as a binary random variable. A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline. The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal plane after $t$ time steps are then given by

@@

Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix covering the contralateral hidden state evolution:

@@

where we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.

@@

We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, in analogy to @eq-lambda-matrix:

@@

Using this, we can again write the joint distribution of ipsi- and contralateral involvement - now also for the case of mid-sagittal extension - algebraically as before in @eq-bilateral-marginal-algebra:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
$$

@@

The terms $P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)$ and $P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)$ are defined solely by sensitivity and specificity of the diagnostic modality. Terms like these already appeared in the definition of the observation matrx in @eq-obs-matrix. The *prior* $P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.

It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $(\mathbf{X}^\text{i}, \mathbf{X}^\text{c})$, but also for e.g. individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all ipsilateral states $\boldsymbol{\xi}^\text{i}_i$ and all contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$. Formally:

$$
P \big( \text{cIV} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
$$ {#eq-marg-over-posterior}

@@

## Involvement Data Consensus {#sec-methods-consensus}

Rather than providing the model with a patient's diagnoses from multiple modalities with different sensitivities and specificities (e.g. PET-CT *and* MRI), we combined them into a single "consensus" diagnosis before inference. This approach was necessary because literature-reported diagnostic imaging sensitivities and specificities did not plausibly match our observations. For instance, in the USZ cohort, 78% of OPSCC patients were diagnosed with ipsilateral LNL II involvement -— a finding difficult to reconcile with typical imaging sensitivities around 80% and specificities below 100% [@debondt_detection_2007;@kyzas_18ffluorodeoxyglucose_2008].

@@

## Computing the Observed and Predicted Prevalence of Involvement Patterns {#sec-prevalence}

We evaluate the model's ability to approximate the observed distribution of lymphatic involvement patterns. We compare the prevalence of selected involvement patterns in the data to the model's predicted prevalence, given patient scenarios.

A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal plane, i.e. $\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$. An involvement pattern specifies all ipsi- and contralateral LNLs' status as "healthy", "involved", or "masked" (ignored).

For example, we may assess contralateral LNL II involvement prevalence under early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$). In the data, {{< var data.early_lateral_cII_total >}} such patients were observed, with {{< var data.early_lateral_cII_match >}} exhibiting contralateral LNL II involvement -- a prevalence of {{< var data.early_lateral_cII_percent >}}%.

To visualize the data prevalence, we plot a *beta posterior* over the true prevalence $q$. It is the result of a uniform beta prior over $q$, multiplied with the binomial likelihood for $k$ out of $n$ patients, given $q$. The resulting distribution has its maximum at $q=k / n$ and nicely captures the uncertainty in the observed cohort: if only 3 out of 10 patients are observed with a specific pattern, the distribution is much wider than if we observe 300 out of 1000.

@@

$$
P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) = \frac{P \left( \text{II}^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{P \left( \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
$$

In the enumator, we marginalize over all ipsi- and contralateral LNLs' involvements, except for LNL II contralaterally. This is similar to the marginalization in @eq-marg-over-posterior, although we are summing over different quantities. In the denominator, we can simply insert the joint distribution over midline extension and diagnose time $P \left( \epsilon, t \right)$ marginalized over $t$ using the early T-category's time-prior.

@@

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters. The bilateral model mostly reproduces the ipsilateral spread parameter values reported in the earlier publication on the unilateral model [@ludwig_modelling_2023]. Any discrepancies may be due to differences in the patient cohorts. Therefore, we omit the analysis of the ipsilateral spread patterns and focus instead on the analysis of contralateral involvement. Lastly, note the relatively small contralateral spread parameters $b^c_v$ compared to the ipsilateral parameters $b^i_v$. This accurately reflects the low prevalence of contralateral lymph node involvement for lateralized tumors.

@@

The framework models the joint probability distribution of midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. This is visualized in @fig-model-algebra-equation, which represents the calculation defined in @eq-bilateral-marginal-algebra. To simplify interpretation, the example focuses only on LNLs II, III, and IV, reducing the state space to $2^3 = 16$ possible states per side, and $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII are excluded, along with their spread parameters, while remaining parameters are set to their mean values from @tbl-midline-params.

![Visual representation of @eq-bilateral-marginal-algebra for midline extension. The left and right matrices represent the time evolution of hidden states for the ipsi- and contralateral necks, respectively. The right matrices distinguish between the cases of no midline extension (top) and midline extension (bottom). The central diagonal matrix shows the time-prior for late T-category tumors. This computation yields the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist.](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}

The left matrix in @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement states, starting from the healthy state $[0,0,0]$. The two right matrices show the contralateral state evolution, distinguishing between the cases of midline extension and no midline extension. At $t=0$, the contralateral neck begins in the healthy state $[0,0,0]$ without midline extension. The central matrix shows the time prior for late T-category tumors. The matrix multiplication results in the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$, visualized in @fig-model-state-dist.

>>>>

In head and neck squamous cell carcinomas (HNSCC) treatments with radiotherapy or surgery, both the primary tumor and clinically detected lymph node metastases are targeted. In addition, current guidelines include large portions of the neck in the elective clinical target volume (CTV-N) [@gregoire_ctbased_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensitymodulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009] to mitigate the risk of regional recurrences from untreated microscopic disease undetectable by in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, this approach must balance minimizing the risk of occult disease in the lymphatic drainage region against the toxicity of unnecessarily irradiating healthy tissue.

These CTV-N guidelines rely on anatomically defined lymph node levels (LNLs) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases within these levels. They often recommend extensive irradiation of both sides of the neck. However, the general prevalence of metastasis in a given LNL does not correspond to an individual patient's risk of occult disease in that region, which depends on their specific state of tumor progression. For example, a patient with no clinically detectable nodal disease (cN0) who has a small, clearly lateralized T1 tumor would receive the same contralateral CTV-N as a patient with significant ipsilateral nodal involvement and an advanced tumor crossing the mid-sagittal plane. Both patients receive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].

To better quantify individualized risk of occult disease, we previously developed an intuitive probabilistic hidden Markov model (HMM) [@ludwig_hidden_2021;@ludwig_modelling_2023], originally based on a conceptually similar a Bayesian network model [@pouymayou_bayesian_2019]. However, these models have been limited to predicting ipsilateral nodal involvement. This work extends the model to include contralateral risk predictions, enabling more personalized radiation volume recommendations for the contralateral neck. By identifying patients with low contralateral risk, the model could guide reductions in the contralateral CTV-N, thereby decreasing radiation-induced toxicity and improving quality of life.

@@

```{python}
#| echo: false
#| label: tbl-data-overview
#| tbl-cap: Overview over the four datasets from four different institutions used to train and evaluate our model. Here, we briefly characterize the total number of OPSCC patients from the respective institution, their median age, what proportion received neck dissection, the N0 portion of patients, what percentage presented with early T-category (T1/T2), and the prevalence of primary tumor midline extension. For a much more detailed look at the data, visit [lyprox.org](https://lyprox.org).
from typing import Literal

@@

The datasets enable analysis of correlations between contralateral LNL involvement and key risk factors. In @fig-data-strat, we illustrate the prevalence of contralateral LNL involvement, stratified by T-category, the number of ipsilaterally involved LNLs, and whether the tumor extends across the mid-sagittal plane.

@@

1. **Midline Extension:** Tumors extending across the mid-sagittal plane should result in a significantly higher probability of contralateral metastases.
2. **T-Category:** Advanced T-category should correspond to an increased risk of nodal disease. In the hidden Markov model this can be modeled using the expected time of diagnosis, as demonstrated previously for the ipsilateral model [@ludwig_hidden_2021].
3. **Ipsilateral Involvement:** The model should be able to capture the correlation between the extent of ipsi- and contralateral involvement. I.e., a more severe ipsilateral involvement should indicate a higher risk for contralateral metastases.

@@

Here, $V$ is the number of LNLs the model considers. The values a LNL may take on are $X_v[t] = 0$ (`False`), meaning the LNL $v$ is free of metastatic disease, or $X_v[t] = 1$ (`True`), corresponding to the presence of clinically detected metastases (i.e., occult or macroscopic disease). In total, there are $2^V$ distinct possible lymphatic involvement patterns, which we enumerate from $\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}$ to $\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}$. Each LNL's state is observed via another binary random variable $Z_v$ that describes the clinical involvement of a LNL based on imaging: $Z_v = 0$ (`False`) indicates that the LNL $v$ is healthy based on clinical diagnosis, and $Z_v = 1$ (`True`) indicates that LNL $v$ was classified as involved. $X_v$ and $Z_v$ are connected through the sensitivity and specificity of the diagnositc modality.

@@

Let us now consider the probability distribution over all possible hidden states $\mathbf{X}[t]$ at time $t$. We can get to this distribution by evolving the healthy starting state $\mathbf{X}[t=0] = \boldsymbol{\xi}_0$ at time $t=0$ step by step, by successively multiplying this vector with the transition matrix $\mathbf{A}$: $\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}$. For later use, we define at this point a matrix $\boldsymbol{\Lambda}$ that collects these distributions for all considered times:

@@

where the $k$-th row in this matrix corresponds to the probability distribution over hidden states after $t=k-1$ time-steps.

@@

The probability distribution over $\mathbf{X}$ for the example of an early T-category patient is given by

@@

The probabiltiy for a patient to present with a diagnosis $\mathbf{Z} = \boldsymbol{\zeta}_i$ a T-category $\text{T}x$ tumor can now be written as:

@@

here, $\left[ \ldots \right]_i$ we denote the $i$-th component of the vector in the square brackets. Note that it is also possible to account for missing involvement information: If a diagnosis (like fine needle aspiration (FNA)) is only available for a subset of all LNLs, we can sum over all those possible complete observed states $\boldsymbol{\zeta}_j$ that match the provided diagnosis.

The term above represents a single patient's contribution to the overall likelihood function that is a product of such terms for each patient. This single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, the parameter space of the unilateral model is:

@@

A straightforward approach to modeling contralateral lymphatic spread would be to use two independent unilateral models, as described in @sec-unilateral, possibly with shared parameters such as the distribution of diagnosis times or spread between LNLs ($t_{vr}$). However, this method would fail to capture the correlation between ipsilateral and contralateral involvement discussed in @sec-data-strat, particularly the observed increase in contralateral involvement with greater severity of ipsilateral spread.

@@

@@

The matrices $\boldsymbol{\Lambda}_\text{i}$ and $\boldsymbol{\Lambda}_\text{c}$ could, in principle, be parameterized with entirely separate parameters, allowing ipsilateral and contralateral spread rates to differ substantially. However, we simplify the parameter space by sharing parameters between the two sides, based on the following three assumptions:

1. **Shared Graph Structure**: Both ipsilateral and contralateral spread are described by the same graph shown in @fig-full-graph.
2. **Symmetric Spread Among LNLs**: The spread among LNLs is assumed to be the same on both sides, reflecting the symmetric structure of the lymphatic system. Consequently, the spread rates between nodes should also be symmetric. This is formalized as:\
  $$
  t_{rv}^\text{c} = t_{rv}^\text{i}
  $$ {#eq-symmetries}
  for all $v \leq V$ and $r \in \operatorname{pa}(v)$ being all nodes that spread to $v$.
1. **Asymmetric Spread from Tumor**: The probabilities of direct spread from the primary tumor are clearly different for the ipsi- and contralateral neck. In addition, tumor spread to the contralateral side varies depending on whether the tumor crosses the mid-sagittal plane. This would result in three sets of rates for tumor spread to the LNLs: (1) the spread to ipsilateral LNLs $b^\text{i}_v$, (2) the spread to contralateral LNLs as long as the tumor is lateralized $b_v^{\text{c},\epsilon=\texttt{False}}$, (3) the spread to contralateral LNLs when the tumor crosses the midline $b_v^{\text{c},\epsilon=\texttt{True}}$. In this work, however, we chose to define the latter set as a linear mix of ipsilateral tumor spread and contralateral spread in case of a clearly lateralized tumor. Thus, using $\alpha \in [0,1]$ as this mixing parameter, we have
  $$
  b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
  $$ {#eq-mixing}
  In this way, the tumor's midline extension causes the contralateral spread to become more like the spread to the ipsilateral side.

The full parameter space of this model is now:

@@

Most tumors crossing the midline at the time of diagnosis likely began as lateralized tumors that grew over the midline at a later point in time. As a result, the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{True}}$ applies only to a subset of time-steps.

To account for this, we model the tumor's extension over the mid-sagittal plane as an additional binary random variable $\epsilon$. A tumor starts as lateralized, with a finite probability $p_\epsilon$ at each time step of crossing the midline. The overall probabilities of a patient having a clearly lateralized tumor or one extending over the mid-sagittal plane after $t$ time steps are then given by

@@

Using this, it is straightforward to write down the matrix of state distributions for all time-steps, as in @eq-lambda-matrix, covering the joint distribution over the contralateral hidden state and the midline extension:

@@

here, we used the transition matrix $\mathbf{A}_\text{c}^{\epsilon=\texttt{False}}$ that depends on the base spread parameters $b_v^{\text{c},\epsilon=\texttt{False}}$.

@@

We can collect the iteratively computed distributions for the midline extension case to define the matrix over the states given all time-steps, as in @eq-lambda-matrix:

@@

In analogy to @eq-bilateral-marginal-algebra, we can now write the joint distribution of ipsi- and contralateral involvement and midline extension algebraically:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c}^\epsilon \right]_{n,m}
$$ {#eq-midline-marginal-algebra}

@@

The terms $P \left( \boldsymbol{\zeta}^\text{i}_k \mid \mathbf{X}^\text{i} \right)$ and $P \left( \boldsymbol{\zeta}^\text{c}_\ell \mid \mathbf{X}^\text{c} \right)$ are defined solely by sensitivity and specificity of the diagnostic modality. These terms already appeared in the definition of the observation matrx in @eq-obs-matrix. The *prior* $P \big( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \boldsymbol{\hat{\theta}}, \text{T}x \big)$ in the above equation is the crucial term that is supplied by a trained model and its parameters $\boldsymbol{\hat{\theta}}$.

It is possible to compute this *posterior* probability of true involvement not only for one fully defined state $(\mathbf{X}^\text{i}, \mathbf{X}^\text{c})$, but also for e.g. individual LNLs: For example, the risk for involvement in the contralateral level IV would be a marginalization over all combination of ipsi- states $\boldsymbol{\xi}^\text{i}_i$ contralateral states $\boldsymbol{\xi}^\text{c}_j$ where $\xi^\text{c}_{j4}=1$. Formally:

$$
P \big( \text{IV}^\text{c} \mid \mathbf{Z}^\text{i}=\boldsymbol{\zeta}^\text{i}_k, \mathbf{Z}^\text{c}=\boldsymbol{\zeta}^\text{c}_\ell, \boldsymbol{\hat{\theta}}, \text{T}x \big) = \sum_k \sum_{\ell \, : \, \xi_{\ell 4}=1} P \big( \mathbf{X}^\text{i} = \boldsymbol{\xi}^\text{i}_k, \mathbf{X}^\text{c} = \boldsymbol{\xi}^\text{c}_\ell \mid \boldsymbol{\zeta}^\text{i}_k, \boldsymbol{\zeta}^\text{c}_\ell, \epsilon, \boldsymbol{\hat{\theta}}, \text{T}x \big)
$$ {#eq-marg-over-posterior}

@@

## Training Data {#sec-training-data}

We trained the model using the dataset of {{< var data.num_patients >}} patients described in @sec-data. The consensus decision on lymphatic involvement is assumed to correspond to the true hidden state of involvement $\mathbf{X}$. Patients with T1 and T2 category tumors have been grouped into an "early" T-category group, those with T3 and T4 tumors into the "advanced" T-category group.

@@

## Comparing the Observed and Predicted Prevalence of Involvement Patterns {#sec-prevalence}

We evaluate the model's ability to describe the observed frequencies of lymphatic involvement patterns. We compare the prevalence of selected involvement patterns in the data to the model's predicted prevalence, given patient scenarios. A "scenario" includes the patient's T-category $\text{T}x$ and whether the tumor extended over the mid-sagittal plane, i.e. $\epsilon=\texttt{True}$ or $\epsilon=\texttt{False}$. An involvement pattern specifies all ipsi- and contralateral LNLs' status as "healthy", "involved", or "masked" (ignored).

For example, in @fig-model-prevalences-overall (top left panel) we assess contralateral LNL II involvement prevalence for early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$). In the data, $n=$ {{< var data.early_lateral_cII_total >}} such patients were observed, with $k=$ {{< var data.early_lateral_cII_match >}} exhibiting contralateral LNL II involvement -- an observed prevalence of $q=$ {{< var data.early_lateral_cII_percent >}}%. To visualize the data prevalence, we plot a *beta posterior* over $q$ -- with a uniform beta prior -- multiplied with the binomial likelihood for $k$ out of $n$ patients, given $q$. The resulting distribution has its maximum at $q=k / n$ and nicely captures the statistical uncertainty in the observed cohort: In the top right panel of @fig-model-prevalences-overall, we consider the case of early T-category tumors extending over the midline. The dataset contains only {{< var data.early_with_midext_cII_total >}} such patients, out of which {{< var data.early_with_midext_cII_match >}} had contralateral LNL II involvement. Consequently, the beta distribution over the observed prevalence is much wider.

@@

$$
\begin{multlined}
P \left( \text{II}^\text{c} \mid \epsilon=\texttt{False}, \text{T}x=\text{early} \right) \\ = \frac{ \sum_k \sum_{\ell \, : \, \xi_{\ell 2}=1}  P \left( \mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}{ \sum_k \sum_\ell P \left(\mathbf{X}^\text{i}=\boldsymbol{\xi}_k^\text{i}, \mathbf{X}^\text{c}=\boldsymbol{\xi}_\ell^\text{c}, \epsilon=\texttt{False} \mid \text{T}x=\text{early} \right)}
\end{multlined}
$$

In the enumator, we marginalize over all combinations of states in both sides of the neck where the contralateral LNL II is involved. This is similar to the marginalization in @eq-marg-over-posterior, although we are summing over different quantities. In the denominator, we simply sum out all LNL involvement, leaving only the joint distribution over midline extension and diagnose time $P \left( \epsilon, t \right)$ marginalized over $t$ using the early T-category's time-prior.

@@

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters. The bilateral model mostly reproduces the ipsilateral spread parameter values reported in the earlier publication on the unilateral model [@ludwig_modelling_2023]. Any discrepancies may be due to differences in the patient cohorts. Therefore, we omit the analysis of the ipsilateral spread patterns and focus instead on the analysis of contralateral involvement. The small contralateral spread parameters $b^c_v$ compared to the ipsilateral parameters $b^i_v$ adequately reflect the low prevalence of contralateral lymph node involvement for lateralized tumors. The mixing parameter $\alpha=$ {{< var model.params.mixing >}} describes that the probability of contralateral spread is higher for tumors extending over the mid-sagittal plane, but still lower than the probability for ipsilateral spread.

The high value of $t_{23}$ ({{< var model.params.IItoIII_spread >}}) compared to the value of $b_3$ ({{< var model.params.ipsi_TtoII_spread >}}) shows that the ipsilateral LNL III is rarely involved without the upstream involvement of LNL II.

@@

The framework models the joint probability distribution of midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. This is visualized in @fig-model-algebra-equation, which represents the calculation defined in @eq-midline-marginal-algebra. To simplify the interpretation, the example focuses only on LNLs II, III, and IV, reducing the state space to $2^3 = 8$ possible states per side, and $2 \times 8 \times 8 = 128$ states in total. LNLs I, V, and VII are excluded, along with their spread parameters, while remaining parameters are set to their mean values from @tbl-midline-params.

![Visual representation of @eq-midline-marginal-algebra. The left and right matrices represent the time evolution of hidden states for the ipsi- and contralateral necks, respectively. The right matrices distinguish between the cases of no midline extension (top) and midline extension (bottom). The central diagonal matrix shows the time-prior for late T-category tumors. This computation yields the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, visualized in @fig-model-state-dist.](figures/fig_model_algebra_equation.svg){#fig-model-algebra-equation}

The left matrix in @fig-model-algebra-equation shows the time evolution of the probability distribution over the ipsilateral involvement states, starting from the healthy state $[0,0,0]$. The two right matrices show the contralateral state evolution, distinguishing between the cases of midline extension and no midline extension. At $t=0$, the contralateral neck begins in the healthy state $[0,0,0]$ without midline extension. The central matrix shows the time prior for late T-category tumors. The matrix multiplication results in the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)$, visualized in @fig-model-state-dist.

This joint distribution is presented as two heatmaps, corresponding to the two states of midline extension. The most likely state is a lateralized tumor with ipsilateral level II involvement and no contralateral involvement, having a probability of approximately 25%. The next most probable state is a lateralized tumor with ipsilateral levels II and III involved, but without contralateral involvement. The most likely state with contralateral involvement corresponds to tumors with midline extension, showing involvement of contralateral level II and ipsilateral levels II and III.
